---
title: >
  Supplementary Information for the `EMMA` package
subtitle: >
  Bioinformatics Group, IMBEI, University Medical Center Mainz
author:
- name: Najla Abassi
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  email: abassina@uni-mainz.de
  orcid: 0000-0001-8357-0938
- name: Annekathrin Nedwed
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  email: anneludt@uni-mainz.de
- name: Federico Marini
  orcid: 0000-0003-3252-7758
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Research Center for Immunotherapy (FZI), Mainz
  email: marinif@uni-mainz.de
date: "`r format(Sys.Date(), '%d %B %Y')`"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: 
    - yeti
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---


```{r knitr, include = FALSE}
knitr::opts_chunk$set(
  comment   = "#>",
  error     = FALSE,
  warning   = FALSE,
  eval      = TRUE,
  message   = FALSE
)
```

In this document, we demonstrate a concrete example of how
`r BiocStyle::Biocpkg("EMMA")` supports transparent reporting and reproducibility of Functional Enrichment Analysis (FEA). We first perform and document enrichment analyses, freeze the corresponding software environment, and subsequently restore this environment in a clean installation to rerun the analyses using the provenance recorded by `EMMA`.

In this example we are using a publicly available dataset from Alasoo et al. "Shared genetic effects on chromatin and gene expression
indicate a role for enhancer priming in immune response", 
published in Nature Genetics, January 2018 
[@Alasoo2018]. The data is made available via the `r BiocStyle::Biocpkg("macrophage")`
Bioconductor package, which contains the files output from the Salmon
quantification (version 0.12.0, with GENCODE v29 reference), as well as the
values summarized at the gene level, which we will use to exemplify.

In the `macrophage` experimental setting, the samples are available from 6
different donors, in 4 different conditions (naive, treated with Interferon
gamma, with SL1344, or with a combination of Interferon gamma and SL1344).

Let's start by loading all the necessary packages:

```{r loadlibraries}
library("EMMA")
library("macrophage")
library("DESeq2")
library("org.Hs.eg.db")
library("clusterProfiler")
```

# Original analysis and environment capture

For demonstration purposes, we will perform Differential Expression
Analysis using the `r BiocStyle::Biocpkg("DESeq2")` framework. 

## Getting a list of Differentially Expressed Genes

```{r load_data}
# load data
data(gse, package = "macrophage")
```

```{r dds_setup}
# set up design
dds_macrophage <- DESeqDataSet(gse, design = ~ line + condition)
# preprocess
rownames(dds_macrophage) <- substr(rownames(dds_macrophage), 1, 15)
keep <- rowSums(counts(dds_macrophage) >= 10) >= 6

dds_macrophage <- dds_macrophage[keep, ]

dds_macrophage
```

```{r differential_expression}
# run DESeq
dds_macrophage <- DESeq(dds_macrophage)
# contrasts
resultsNames(dds_macrophage)
```

```{r get_DE_results}
# get de res for 1st contrast
IFNg_vs_naive <- results(dds_macrophage,
                         contrast = c("condition", "IFNg", "naive"),
                         lfcThreshold = 1, alpha = 0.05)
IFNg_vs_naive <- lfcShrink(dds_macrophage, coef = "condition_IFNg_vs_naive",
                           res = IFNg_vs_naive,
                           type = "apeglm")
IFNg_vs_naive$SYMBOL <- rowData(dds_macrophage)$SYMBOL

# get de res for 2nd contrast
SL1344_vs_naive <- results(dds_macrophage,
                         contrast = c("condition", "SL1344", "naive"),
                         lfcThreshold = 1, alpha = 0.05)
SL1344_vs_naive <- lfcShrink(dds_macrophage, coef = "condition_SL1344_vs_naive",
                           res = SL1344_vs_naive,
                           type = "apeglm")
SL1344_vs_naive$SYMBOL <- rowData(dds_macrophage)$SYMBOL

# get de res for 3rd contrast
IFNg_SL1344_vs_naive <- results(dds_macrophage,
                         contrast = c("condition", "IFNg_SL1344", "naive"),
                         lfcThreshold = 1, alpha = 0.05)
IFNg_SL1344_vs_naive <- lfcShrink(dds_macrophage, coef = "condition_IFNg_SL1344_vs_naive",
                           res = IFNg_SL1344_vs_naive,
                           type = "apeglm")
IFNg_SL1344_vs_naive$SYMBOL <- rowData(dds_macrophage)$SYMBOL

# sort by adjusted p value
de_list <- list(
  IFNg_vs_naive = IFNg_vs_naive,
  SL1344_vs_naive = SL1344_vs_naive,
  IFNg_SL1344_vs_naive = IFNg_SL1344_vs_naive
)

de_list <- lapply(de_list, function(df) {
  df <- df[order(df$padj), ]
  df <- df[!is.na(df$padj) & df$padj <= 0.05, ]
  df
})
```

```{r bg_genes}
# set background gene list
gene_universe <- rownames(dds_macrophage)
```

## Performing Functional Enrichment Analysis (FEA) with `r BiocStyle::Biocpkg("EMMA")`

Now that we have a list of DE genes for different contrasts, we can perform Functional
Enrichment Analysis.

### FEA with `clusterProfiler::enrichGO()`

In the following example, we will use the function `enrichGO()`

```{r enrichGO}
# 1st contrast
de_res1 <- de_list$IFNg_vs_naive

fea_res <- enrichGO(gene = rownames(de_res1),
                    universe = gene_universe,
                    keyType = "ENSEMBL",
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.05,
                    qvalueCutoff = 0.1) |> 
  EMMA_run(args_form = "evaluated",
           store_session_info = TRUE)
```

```{r summry_record_GO}
EMMA_show(fea_res)
```

### FEA with `clusterProfiler::enrichKEGG()`

For the following contrast, we will use the function `enrichKEGG()`

```{r enrichKEGG}
# 2nd contrast
de_res2 <- de_list$IFNg_SL1344_vs_naive

# convert ENSEMBL to ENTREZID
universe_map <- mapIds(
    org.Hs.eg.db,
    keys = gene_universe,
    keytype = "ENSEMBL",
    column = "ENTREZID",
    multiVals = "first"
)

universe_entrez <- unique(na.omit(universe_map))
genes_entrez <- universe_map[rownames(de_res2)]
genes_entrez <- unique(na.omit(genes_entrez))

fea_res2 <- enrichKEGG(gene = genes_entrez,
                       universe = universe_entrez,
                       keyType = "ncbi-geneid",
                       organism = "hsa",
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.1) |> 
  EMMA_run(args_form = "evaluated",
           store_session_info = TRUE)
```

```{r summry_record_KEGG}
EMMA_show(fea_res2)
```

Here no KEGG version is recorded because this information is not exposed by the underlying remote resource.

```{r fea_res_list}
fea_list <- list(IFNg_vs_naive = fea_res,
                 IFNg_SL1344_vs_naive = fea_res2)
```

## Save analysis results

To save all the generated results, we will construct a DeeDeeExperiment object, to keep all analysis elements together. This step is optional.

```{r dde}
library(DeeDeeExperiment)

dde <- DeeDeeExperiment(sce = dds_macrophage,
                        de_results = de_list,
                        enrich_results = fea_list)
dde
```

```{r save, eval=FALSE}
saveRDS(dde, "dde_macrophage.RDS")
```

## Capture the analysis environment with `EMMA_freeze()`

```{r emma_freeze}
project_path <- getwd()
EMMA_freeze(project = project_path,
            file = "EMMA_analysis.lock",
            pkgs = loadedNamespaces(),
            prompt = FALSE,
            force = TRUE)
```

# Environment restoration and reproducibility check

To evaluate environment restoration independently of the original installation, the analysis was reproduced in a Docker container based on R 4.6.1. 

From within the Docker container, we recreate the analysis result using the following command:

```{reval=FALSE}
library("renv")
restore(lockfile = "/home/rstudio/emma_supplement/EMMA_analysis.lock")
```

Then we load the analysis results

```{r load_dde, eval=FALSE}
library("DeeDeeExperiment")
library("EMMA")
library("org.Hs.eg.db")
library("clusterProfiler")

dde <- readRDS("/home/rstudio/emma_supplement/dde_macrophage.RDS")

# check
dde
```

We extract the FEA results as well as the **provenance record**, i.e. `EMMA_record`

```{r getFEA, eval=FALSE}
fea_res1 <- getFEA(dde, fea_name = "IFNg_vs_naive", format = "original")
fea_res2 <- getFEA(dde, fea_name = "IFNg_SL1344_vs_naive", format = "original")
```

```{r get_record, eval=FALSE}
rec1 <- EMMA_get_record(fea_res1)
rec2 <- EMMA_get_record(fea_res2)
```

## Reproducing the GO enrichment analysis

```{r get_input, eval=FALSE}
rec1$method$call

gene_list <- rec1$input$arguments$gene
background <- rec1$input$arguments$universe
```

```{r run_fea_go, eval=FALSE}
fea_res_new <- enrichGO(gene = gene_list,
                        universe = background,
                        keyType = "ENSEMBL",
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 0.05,
                        qvalueCutoff = 0.1) |> 
  EMMA_run(args_form = "evaluated",
           store_session_info = TRUE)
```

```{r compare_1, eval=FALSE}
# compare the FEA results
all.equal(as.data.frame(fea_res1),
          as.data.frame(fea_res_new))
```

## Reproducing the KEGG enrichment analysis

```{r get_input2, eval=FALSE}
rec2$method$call

gene_list2 <- rec2$input$arguments$gene
background2 <- rec2$input$arguments$universe
```

```{r run_fea_kegg, eval=FALSE}
fea_res_new2 <- enrichKEGG(gene = gene_list2,
                          universe = background2,
                          keyType = "ncbi-geneid",
                          organism = "hsa",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.1) |> 
  EMMA_run(args_form = "evaluated",
           store_session_info = TRUE)
```

```{r compare_2, eval=FALSE}
# compare the FEA results
all.equal(as.data.frame(fea_res2),
          as.data.frame(fea_res_new2))
```

# Session info {.unnumbered}

```{css}
/*| echo: false */

.smaller {
  font-size: 10px
}
```

```{r sessioinfo}
#| class-output: smaller

sessionInfo()
```

# References {.unnumbered}
